Manuscript progress report

### 2023-12-18 Docs/manuscript.Rmd
example of issue with debias step in NHNES

---
title: "Manuscript progress report"
author: "Ying Jin"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
    keep_tex: yes
  html_document:
    self_contained: no
    number_sections: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float: yes
    font: 12pt
    keep_tex: yes
---

@@

fig_fgfpca_subset <- pred_subset_fGFPCA[[1]] %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "fGFPCA", col = "Maximum observation time")+
  scale_x_continuous(breaks = seq(0, 1, by = 0.2))

fig_adglmm_subset <- pred_subset_adglmm[[1]] %>% 
  filter(id %in% rand_id2) %>%
  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "GLMMadaptive", col = "Maximum observation time")+
  scale_x_continuous(breaks = seq(0, 1, by = 0.2))

@@

We take 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction.

In the GLMMadaptive model, I could only fit a model with random intercept. It took more than 20 minutes for model fitting.

@@

```{r small_exp}
# a smaller example of binary functions
df_nhanes %>% select(SEQN, Z, sind) %>% 
  filter(SEQN %in% rand_id3 & Z == 1) %>% 
  mutate(SEQN = as.factor(SEQN)) %>% 
  ggplot()+
  geom_point(aes(x=sind, y = SEQN), size = 0.5)+
  labs(x="Time", y="Subject", title = "NHANES binary activity indicator")
```

@@

>>>>

---
title: "Manuscript progress report"
author: "Ying Jin"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    number_sections: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float: yes
    font: 12pt
    keep_tex: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
    keep_tex: yes
---

@@

fig_fgfpca_subset <- pred_subset_fGFPCA[[1]] %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "fGFPCA", col = "Maximum observation time")+
  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
  theme(axis.text = element_text(size = 5))

fig_adglmm_subset <- pred_subset_adglmm[[1]] %>% 
  filter(id %in% rand_id2) %>%
  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "GLMMadaptive", col = "Maximum observation time")+
  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
  theme(axis.text = element_text(size = 5))
```

@@

We take 500 subjects for training, 500 for out-of-sample prediction. When using the full sample The debiase step (mgcv::bam) in fGFPCA took a really a long time, and GLMM adaptive only works for intercept-only model. In the debias step, re-evaluated eigenvalues also lost their decreasing nature.

In addition to reduce sample size, maybe we could just drop the debias step and only interpolate the eigenfunctions?

@@

@@

## Example of continuous and binary NHANES

```{r}
df_nhanes_cont <- readRDS(here("Data/NHANES_AC_processed.rds"))
p_cont <- df_nhanes_cont %>% 
  filter(SEQN %in% sample(unique(df_nhanes_cont$SEQN), 4) & WEEKDAY==6) %>% 
  select(starts_with("min"), SEQN) %>%
  pivot_longer(1:1440, names_to = "Minute") %>% 
  mutate(Minute = gsub("MIN", "", Minute)) %>%
  mutate(Minute = as.numeric(Minute),
         SEQN=as.factor(SEQN)) %>%
  ggplot(aes(x=Minute, y=value, col=SEQN))+
  geom_point(size = 0.5, alpha=0.5)+
  geom_smooth(se=F, size = 0.5)+
  scale_x_continuous(breaks = seq(0, 1440, by = 480),
                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
  theme(axis.text = element_text(size = 5))+
  labs(y="", title = "NHANES Activitiy count (2003-2004)")
```

```{r}
# a example of continuous and binary NHANES data
p_bi <- df_nhanes %>% select(SEQN, Z, sind) %>% 
  filter(SEQN %in% rand_id3 & Z == 1) %>% 
  mutate(SEQN = as.factor(SEQN)) %>% 
  ggplot()+
  geom_point(aes(x=sind, y = SEQN), size = 0.5)+
  scale_x_continuous(breaks = seq(0, 1440, by = 480),
                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
  theme(axis.text = element_text(size = 5))+
  labs(x="Time", y="Subject", title = "NHANES binary activity indicator (2011-2014)")
```

```{r data_exp, fig.height=4, fig.width=10}
ggarrange(p_cont, p_bi, nrow = 1)
```

## Dynamic prediction example

```{r dynpred_exp,  fig.height=4, fig.width=10}
ggarrange(
  pred_list_all[[1]] %>% 
    filter(id==501) %>%
    mutate(eta_i = ifelse(t<=0.2, eta_i, NA)) %>%
    ggplot()+
    geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+
    geom_line(aes(x=t, y=pred0.2, col = "Predicted"), na.rm = T, 
              linetype = "dashed")+
    geom_vline(xintercept = 0.2, linetype = "dashed")+
    labs(x="Time", y="")+
    scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
    theme(axis.text = element_text(size = 5))+
    ylim(-1.2, 1.7),

 pred_list_all[[1]] %>% 
    filter(id==501) %>%
    mutate(eta_i = ifelse(t<=0.4, eta_i, NA)) %>%
    ggplot()+
    geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+
    geom_line(aes(x=t, y=pred0.4, col = "Predicted"), na.rm = T, 
              linetype = "dashed")+
    geom_vline(xintercept = 0.4, linetype = "dashed")+
    labs(x="Time", y="")+
    scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
    theme(axis.text = element_text(size = 5))+
   ylim(-1.2, 1.7),
 nrow = 1, common.legend = T
  )
